Forecasting the 2024 Election: Tight Margins and Swing State Uncertainty

### 2024-11-05 paper/paper.qmd
finish

The rest of this paper is structured as follows:  @sec-mydatasection presents the data sources and methodology, @sec-mymodel presents the forecasting model, then  @sec-myresults followed by  @sec-mydisc, which discusses the results and their implications.  @sec-mylimits concludes with a discussion of the limitations of the study and suggestions for future research, followed by @sec-appenx with The Washington Post’s pollster methodology and a budgeted idealized methodology.

@@

## Outcome Variable
### Distribution of Polls and State-Level Candidate Preference
The chosen candidate—whether a poll favors Harris or Trump—serves as the outcome variable in this analysis. [@fig-candidate] shows the distribution of polls between the two candidates, Harris and Trump, revealing a nearly equal split in support, with 49.8% favoring Trump and 50.2% favoring Harris. This close distribution highlights a competitive polling landscape, suggesting that public opinion is almost evenly divided between the candidates.

In [@fig-candbystate], the state-level distribution of support further emphasizes this balance, with only slight variations in candidate preference across states. For instance, states like Maryland, Montana, Nevada, Ohio, and Utah show a marginally higher level of support for Harris, representing the Democratic party. Conversely, in states such as Florida, Minnesota, and Pennsylvania, there is slightly greater support for Trump, representing the Republican party. These minor differences indicate regional trends but reinforce the overall theme of a closely contested race.
```{r}
#| label: fig-candidate
#| fig-cap: Distribution of Polls by Harris vs Trump
#| echo: false
#| warning: false
#| fig-align: center
ggplot(clean_president_polls, aes(x = factor(candidate_chosen), fill = factor(candidate_chosen))) +
  geom_bar(alpha = 0.7) +
  labs(x = "Chosen Candidate", 
       y = "Number of Polls",
       fill = "Chosen Candidate") +  # This sets the legend title
  scale_x_discrete(labels = c("0" = "Trump", "1" = "Harris")) +
  theme_classic() +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), labels = c("Trump", "Harris")) +
  geom_text(stat = "count", aes(label = scales::percent(..count../sum(..count..), 
                                                        accuracy = 0.2)), vjust = -0.5) 
```

```{r}
#| label: fig-candbystate
#| tbl-cap: Distribution of Candidate Preference by State
#| echo: false
#| warning: false
#| fig-align: center
ggplot(clean_president_polls, aes(x = state, fill = factor(candidate_chosen, labels = c("Trump", "Harris")))) +
  geom_bar(position = "fill") +
  labs(x = "State", y = "Proportion", fill = "Chosen Candidate") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, margin = margin(t = 5), size = 8)
  )
```

@@

The prevalence of smaller sample sizes suggests greater variability in the precision of poll estimates, as smaller samples generally introduce more uncertainty. In contrast, larger sample sizes provide more stable and reliable reflections of public sentiment. This distribution emphasizes the need to account for sample size in the analysis to ensure that more reliable, larger polls are given appropriate weight, thus enhancing the accuracy of public opinion estimates.
```{r}
#| label: fig-sample
#| fig-cap: Distribution of Sample Size
#| echo: false
#| warning: false
#| fig-align: center
ggplot(clean_president_polls, aes(x = sample_size)) +
  geom_histogram(binwidth = 200, fill = "darkred", color = "black", alpha = 0.7) +
  labs(x = "Sample Size", y = "Frequency") +
  theme_classic()
```

@@

A few outliers in both distributions show support above 60% or below 40%, representing occasional strong deviations, though these instances are rare. This pattern reflects a balanced public sentiment with minor variation between the candidates and points to a highly competitive race. The symmetry and overlap of the distributions reinforce the close competition between Trump and Harris, indicating an almost evenly divided public opinion across the dataset.
```{r}
#| label: fig-percentage
#| tbl-cap: Distribution of Candidate Support by Chosen Candidate
#| echo: false
#| warning: false
#| fig-align: center
ggplot(clean_president_polls, aes(x = pct, fill = candidate_name)) +
  geom_density(alpha = 0.5) +
  labs(x = "Percentage Support", y = "Density") +
  scale_fill_manual(values = c("blue", "red"), name = "Candidate") +
  theme_classic()
```

@@

These insights are crucial, as they enable the model to assign weights based on pollster reliability, improving the accuracy of the analysis. By prioritizing data from higher-grade pollsters, the model can more effectively capture public sentiment, minimizing the impact of potential biases or inaccuracies from less reliable sources.
```{r}
#| label: fig-numeric
#| fig-cap: Distribution of Numeric Grade
#| echo: false
#| warning: false
#| fig-align: center
ggplot(clean_president_polls, aes(x = numeric_grade)) +
  geom_bar(fill = "darkred", alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 3) +
  labs(x = "Numeric Grade", y = "Frequency") +
  theme_classic()
```

```{r}
#| label: fig-pollsters
#| fig-cap: Distribution of Pollsters by Numeric Grade 
#| echo: false
#| warning: false
#| fig-align: center
# Calculate the number of polls for each pollster with more than 20 polls
pollsters_over_20 <- clean_president_polls %>%
  group_by(pollster) %>%
  filter(n() > 20) %>%
  summarize(
    num_polls = n(),
    avg_numeric_grade = mean(numeric_grade, na.rm = TRUE)
  )

# Plot 
ggplot(pollsters_over_20, aes(x = reorder(pollster, -num_polls), y = num_polls)) +
  geom_bar(stat = "identity", aes(fill = avg_numeric_grade), alpha = 0.7) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Numeric Grade") +
  labs(x = "Pollster", y = "Number of Polls") +
  geom_text(aes(label = round(avg_numeric_grade, 2)), vjust = -0.5, size = 2.5) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

@@

[@fig-support] presents a map indicating the leading candidate by state, with blue for Kamala Harris, red for Donald Trump, and gray for states with insufficient polling data. The map reveals a geographic divide, with Trump leading in many central and southern states, while Harris is favored in parts of the Northeast and West. This distribution aligns with typical U.S. regional political trends, where certain areas have a historical leaning toward one party. The map provides a visual summary of these regional preferences, highlighting how candidate support varies across the country.
```{r}
#| label: fig-numberstate
#| fig-cap: Number of Pollsters by State
#| echo: false
#| warning: false
#| fig-align: center
ggplot(clean_president_polls, aes(x = reorder(state, -table(state)[state]))) +
  geom_bar(fill = "darkblue") +
  labs(x = "State", y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig-support
#| tbl-cap: Candidate with Highest Support Percentage by State
#| echo: false
#| warning: false
#| fig-align: center
# Summarize data to find the candidate with the highest percentage in each state
state_winners <- clean_president_polls %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) %>%
  group_by(state) %>%
  summarize(winner = candidate_name[which.max(pct)])

@@

# Plot the election map
ggplot(data = merged_data, aes(x = long, y = lat, group = group, fill = winner)) +
  geom_polygon(color = "black", size = 0.3) + 
  labs(fill = "Winner") +
  scale_fill_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  coord_fixed(1.3) + 
  labs(title = "Leading Candidate by State") +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )
```

@@

* Component Explanation
- $y_i$: Binary outcome variable indicating whether a respondent supports Harris (1) or not (0). \\
- $\beta_0$:Intercept term representing the baseline log-odds of support for Harris when the percentage support is zero. \\
- $\beta_1$: Coefficient for the percentage support for Harris (Pct), reflecting how changes in support percentage affect the log-odds of selecting Harris. \\
- $u_{\text{pollster}}$:Random effect for the specific pollster, accounting for variability in support across different pollsters. \\
- $u_{\text{state}}$: Random effect for the state in which the poll is conducted, capturing regional variations in support.

@@

### Diagnostics
To validate the Bayesian framework, several diagnostic checks were conducted to assess robustness and ensure that underlying assumptions were satisfied. Key among these were convergence assessments and posterior predictive checks. These checks provided a visual representation of how well the predictions align with the observed data. The graph reveals a close overlap between the observed outcomes (\(y\)) and the replicated predictions (\(y_{\text{rep}}\)), demonstrating effective capture of the distribution of voter support for Harris. This strong alignment suggests that the approach not only fits the training data well but also generalizes effectively to new observations. The flexibility of this framework allows for the incorporation of prior information and consideration of variability due to pollster and state effects, enhancing overall robustness (Refer to the @sec-appc, [@tbl-pp] for a visual illustration).

@@

[@fig-national-vote] shows the national predicted popular vote percentages for Harris (50.44%) and Trump (49.56%), calculated by averaging the mean predictions across all states. This aggregation reveals a very narrow 0.88% lead for Harris, highlighting a highly competitive race. This reveals a lack of dominance in candidacy and general uncertainty of pollster outcomes, suggesting that even this slight lead may not be substantial.
```{r}
#| label: fig-national-vote
#| fig-cap: "National Popular Vote Prediction"
#| echo: false
#| warning: false
#| fig-align: center

# Calculate overall percentage for Harris and Trump
overall_percentages <- data.frame(
  Candidate = c("Harris", "Trump"),
  Percentage = c(overall_percentage_harris, overall_percentage_trump)
)

# Plot national popular vote prediction
ggplot(overall_percentages, aes(x = Candidate, y = Percentage, fill = Candidate)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.2f%%", Percentage)), vjust = -0.5) +
  labs(y = "Percentage", x = "Candidate") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_classic()
```

Additionally, [@fig-heatmap] presents a heat map of Harris's predicted probability by state, illustrating her lead over the majority of states but revealing that this advantage is generally weak, with most not peaking over a 1% advantage. Most states show only a slight preference for Harris over Trump, with the margin of error spanning both candidates in many cases. Specifically, out of the 43 states with available data, 25 states lean toward Harris, yet none of these states shows a substantial lead over more than a few percent over Trump. This pattern additionally shows the substantial closeness of the race, as evidenced by the margin of error, where overlapping intervals highlight the slim advantage and limited certainty in Harris's predicted support.
```{r}
#| label: fig-heatmap
#| fig-cap: "Heatmap of Predicted Support for Harris by State"
#| echo: false
#| warning: false
#| fig-align: center

@@

# Plot heatmap for Harris's support by state, with NA values as dark grey
ggplot(merged_data, aes(x = long, y = lat, group = group, fill = avg_predicted_prob_harris)) +
  geom_polygon(color = "black", size = 0.3) + 
  scale_fill_gradient(low = "lightblue", high = "blue", na.value = "darkgrey", name = "Predicted Support") +
  coord_fixed(1.3) + 
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )
```

[@fig-swing-states] zooms in on the seven swing states, each of which shows a slight lean toward Harris. The graph includes a red dashed line at the 50% threshold, and although Harris is estimated to lead in these swing states, the margin is extremely tight, with most probabilities hovering just above 50%. Only Michigan and North Carolina show a slight lead of just over 1%, reinforcing the sensitivity of these predictions. This pattern suggests that even small shifts in voter sentiment could tip these critical states in either direction.
```{r}
#| label: fig-swing-states
#| fig-cap: "Predicted Probability of Harris Winning by State with Swing State Highlight and 50% Threshold"
#| echo: false
#| warning: false
#| fig-align: center

@@

# Plot with enhancements
ggplot(state_predictions, aes(x = reorder(state, avg_predicted_prob_harris), y = avg_predicted_prob_harris, fill = swing_state)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  scale_fill_manual(values = c("Swing State" = "darkblue", "Non-Swing State" = "lightgrey")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red", size = 0.8) +
  labs(
    title = "Predicted Probability of Harris Winning by State",
    x = "State",
    y = "Predicted Probability for Harris",
    fill = "State Type"
  ) +
  annotate("text", x = 1, y = 0.51, label = "0.5 Threshold (Neutral)", color = "red", hjust = 0) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()
```

Looking closer at the margin of error, in [@fig-margin] we can see Harris’s predicted probabilities by state. States to the right of the dashed line (above 0.5) lean towards Harris, while those to the left lean towards Trump. In swing states like North Carolina, Michigan, and Pennsylvania, Harris’s minimal lead is coupled with error bars that overlap this threshold. This suggests a high level of uncertainty—small shifts could easily change the lead in these states. Harris has a slight edge, but in many states, this lead is within the margin of error, thus showing that the election remains too close to call.
```{r}
#| label: fig-margin
#| fig-cap: "State-Level Predicted Probability for Harris with Margin of Error"
#| echo: false
#| warning: false
#| fig-align: center

# Calculate margin of error for each state
state_margins <- clean_president_polls %>%
  group_by(state) %>%
  summarize(
    avg_predicted_prob_harris = mean(predicted_prob_harris),
    se_predicted_prob_harris = sd(predicted_prob_harris) / sqrt(n()),
    margin_of_error = 1.96 * se_predicted_prob_harris
  ) %>%
  mutate(
    lower_bound = avg_predicted_prob_harris - margin_of_error,
    upper_bound = avg_predicted_prob_harris + margin_of_error
  )

# Plot with error bars
ggplot(state_margins, aes(x = reorder(state, avg_predicted_prob_harris), y = avg_predicted_prob_harris)) +
  geom_point(color = "blue") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2, color = "blue") +
  labs(
    x = "State",
    y = "Predicted Probability for Harris"
  ) +
  coord_flip() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(subtitle = "Dashed line represents a 50% threshold")
```

@@

```{r}
#| label: tbl-summary-mod
#| tbl-cap: "Summary of Bayesian Model"
#| echo: false
#| warning: false
#| tbl-align: center

# Display model summary with customized formatting
modelsummary(bayesian_model, 
             output = "kableExtra")

@@

\appendix

# Appendix {#sec-appenx}

# Appendix A: Methodology Analysis of The Washington Post Polling  
With an evaluation of sampling methodology, recruitment, handling non-response, and questionnaire design, this appendix offers an analysis of the polling methodology used by The Washington Post in collaboration with ABC News. The objective is to analyze these approaches' strengths and weaknesses and determine how they affect polling accuracy.

@@

# Appendix B: Idealized Survey & Methodology - $100K Budget

@@

This survey methodology uses stratified sampling, multi-channel recruitment, and data validation procedures to ensure accurate forecasting of the 2024 US Presidential Election. Through poll-of-polls aggregation, it offers a stable and reliable prediction by mitigating fluctuations in individual polls. The design aims to balance accuracy, inclusivity, and efficiency within a $100K budget, offering meaningful insights into voter behavior and sentiment.

# Appendix C {#sec-appc}
```{r}
#| include: false
#| warning: false
#| message: false
### Model Validation/Checking ###
# Set seed and create train/test split
set.seed(123)
train_indices <- sample(seq_len(nrow(clean_president_polls)), size = 0.7 * nrow(clean_president_polls))
training_data <- clean_president_polls[train_indices, ]
training_data <- training_data %>%
  mutate(pct_scaled = scale(pct))
testing_data <- clean_president_polls[-train_indices, ]

>>>>

The rest of this paper is structured as follows:  @sec-mydatasection presents the data sources and methodology, @sec-mymodel presents the forecasting model, then  @sec-myresults followed by  @sec-mydisc, which discusses the results and their implications.  @sec-mylimits concludes with a discussion of the limitations of the study and suggestions for future research, followed by @sec-appenxA with The Washington Post’s pollster methodology, and @sec-appenxB a budgeted idealized methodology.

@@

## Outcome Variable
The chosen candidate—whether a poll favors Harris or Trump—serves as the outcome variable in this analysis. [@fig-candidate] shows the distribution of polls between the two candidates, Harris and Trump, revealing a nearly equal split in support, with 49.8% favoring Trump and 50.2% favoring Harris. This close distribution highlights a competitive polling landscape, suggesting that public opinion is almost evenly divided between the candidates.

In [@fig-candbystate], the state-level distribution of support further emphasizes this balance, with only slight variations in candidate preference across states. For instance, states like Maryland, Montana, Nevada, Ohio, and Utah show a marginally higher level of support for Harris, representing the Democratic party. Conversely, in states such as Florida, Minnesota, and Pennsylvania, there is slightly greater support for Trump, representing the Republican party. These minor differences indicate regional trends but reinforce the overall theme of a closely contested race.
```{r}
#| label: fig-candidate
#| fig-cap: Distribution of Polls by Harris vs Trump
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
ggplot(clean_president_polls, aes(x = factor(candidate_chosen), fill = factor(candidate_chosen))) +
  geom_bar(alpha = 0.7) +
  labs(x = "Chosen Candidate", 
       y = "Number of Polls",
       fill = "Chosen Candidate") +  # This sets the legend title
  scale_x_discrete(labels = c("0" = "Trump", "1" = "Harris")) +
  theme_classic() +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"), labels = c("Trump", "Harris")) +
  geom_text(stat = "count", aes(label = scales::percent(..count../sum(..count..), 
                                                        accuracy = 0.2)), vjust = -0.5, size = 3)  +
  theme(
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```

```{r}
#| label: fig-candbystate
#| fig-cap: Distribution of Candidate Preference by State
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
ggplot(clean_president_polls, aes(x = state, fill = factor(candidate_chosen, labels = c("Trump", "Harris")))) +
  geom_bar(position = "fill") +
  labs(x = "State", y = "Proportion", fill = "Chosen Candidate") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, margin = margin(t = 5), size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```
\newpage

@@

The prevalence of smaller sample sizes suggests greater variability in the precision of poll estimates, as smaller samples generally introduce more uncertainty. In contrast, larger sample sizes provide more stable and reliable reflections of public sentiment. This distribution emphasizes the need to account for sample size in the analysis to ensure that more reliable, larger polls are given appropriate weight, thus enhancing the accuracy of public opinion estimates.
```{r}
#| label: fig-sample
#| fig-cap: Distribution of Sample Size
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
ggplot(clean_president_polls, aes(x = sample_size)) +
  geom_histogram(binwidth = 200, fill = "darkred", color = "black", alpha = 0.7) +
  labs(x = "Sample Size", y = "Frequency") +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
  )
```

@@

A few outliers in both distributions show support above 60% or below 40%, representing occasional strong deviations, though these instances are rare. This pattern reflects a balanced public sentiment with minor variation between the candidates and points to a highly competitive race. The symmetry and overlap of the distributions reinforce the close competition between Trump and Harris, indicating an almost evenly divided public opinion across the dataset.
```{r}
#| label: fig-percentage
#| fig-cap: Distribution of Candidate Support by Chosen Candidate
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
ggplot(clean_president_polls, aes(x = pct, fill = candidate_name)) +
  geom_density(alpha = 0.5) +
  labs(x = "Percentage Support", y = "Density") +
  scale_fill_manual(values = c("blue", "red"), name = "Candidate") +
  theme_classic() +
  theme(
    axis.text.x = element_text(hjust = 1, size = 4), 
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```

@@

These insights are crucial, as they enable the model to assign weights based on pollster reliability, improving the accuracy of the analysis. By prioritizing data from higher-grade pollsters, the model can more effectively capture public sentiment, minimizing the impact of potential biases or inaccuracies from less reliable sources.
```{r}
#| label: fig-numeric
#| fig-cap: Distribution of Numeric Grade
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
ggplot(clean_president_polls, aes(x = numeric_grade)) +
  geom_bar(fill = "darkred", alpha = 0.7) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 2) +
  labs(x = "Numeric Grade", y = "Frequency") +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
  )
```

```{r}
#| label: fig-pollsters
#| fig-cap: Distribution of Pollsters by Numeric Grade 
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
# Calculate the number of polls for each pollster with more than 20 polls
pollsters_over_20 <- clean_president_polls %>%
  group_by(pollster) %>%
  filter(n() > 30) %>%
  summarize(
    num_polls = n(),
    avg_numeric_grade = mean(numeric_grade, na.rm = TRUE)
  )

# Plot 
ggplot(pollsters_over_20, aes(x = reorder(pollster, -num_polls), y = num_polls)) +
  geom_bar(stat = "identity", aes(fill = avg_numeric_grade), alpha = 0.7) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Numeric Grade") +
  labs(x = "Pollster", y = "Number of Polls") +
  geom_text(aes(label = round(avg_numeric_grade, 2)), vjust = -0.5, size = 2.5) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 4), 
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```
\newpage

@@

[@fig-support] presents a map indicating the leading candidate by state, with blue for Kamala Harris, red for Donald Trump, and gray for states with insufficient polling data. The map reveals a geographic divide, with Trump leading in many central and southern states, while Harris is favored in parts of the Northeast and West. This distribution aligns with typical U.S. regional political trends, where certain areas have a historical leaning toward one party. The map provides a visual summary of these regional preferences, highlighting how candidate support varies across the country.
```{r}
#| label: fig-numberstate
#| fig-cap: Number of Pollsters by State
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
ggplot(clean_president_polls, aes(x = reorder(state, -table(state)[state]))) +
  geom_bar(fill = "darkred") +
  labs(x = "State", y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 4),
        axis.text.y = element_text(hjust = 1, size = 6),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
  )
```

```{r}
#| label: fig-support
#| fig-cap: Candidate with Highest Support Percentage by State
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
# Summarize data to find the candidate with the highest percentage in each state
state_winners <- clean_president_polls %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump")) %>%
  group_by(state) %>%
  summarize(winner = candidate_name[which.max(pct)])

@@

# Plot the election map
ggplot(data = merged_data, aes(x = long, y = lat, group = group, fill = winner)) +
  geom_polygon(color = "black", size = 0.3) + 
  labs(fill = "Winner") +
  scale_fill_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  coord_fixed(1.3) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```
\newpage

@@

* Component Explanation
- $y_i$: Binary outcome variable indicating whether a respondent supports Harris (1) or not (0). 
- $\beta_0$:Intercept term representing the baseline log-odds of support for Harris when the percentage support is zero. 
- $\beta_1$: Coefficient for the percentage support for Harris (Pct), reflecting how changes in support percentage affect the log-odds of selecting Harris. 
- $u_{\text{pollster}}$:Random effect for the specific pollster, accounting for variability in support across different pollsters. 
- $u_{\text{state}}$: Random effect for the state in which the poll is conducted, capturing regional variations in support.

@@

### Diagnostics
To validate the Bayesian framework, several diagnostic checks were conducted to assess robustness and ensure that underlying assumptions were satisfied. Key among these were convergence assessments and posterior predictive checks. These checks provided a visual representation of how well the predictions align with the observed data. The graph reveals a close overlap between the observed outcomes \(y\) and the replicated predictions ($y_{\text{rep}}$), demonstrating effective capture of the distribution of voter support for Harris. This strong alignment suggests that the approach not only fits the training data well but also generalizes effectively to new observations. The flexibility of this framework allows for the incorporation of prior information and consideration of variability due to pollster and state effects, enhancing overall robustness (Refer to the @sec-appc, [@tbl-pp] for a visual illustration).

@@

The analysis of the national predicted popular vote percentages indicates that Kamala Harris is projected to receive 50.44% of the vote, while Donald Trump is expected to garner 49.56%. These figures, illustrated in [@fig-national-vote] and calculated by averaging the mean predictions across all states, reveal a narrow lead of just 0.88% for Harris, highlighting a highly competitive race. This reveals a lack of dominance in candidacy and general uncertainty of pollster outcomes, suggesting that even this slight lead may not be substantial.
```{r}
#| label: fig-national-vote
#| fig-cap: "National Popular Vote Prediction"
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3
# Calculate overall percentage for Harris and Trump
overall_percentages <- data.frame(
  Candidate = c("Harris", "Trump"),
  Percentage = c(overall_percentage_harris, overall_percentage_trump)
)

# Plot national popular vote prediction
ggplot(overall_percentages, aes(x = Candidate, y = Percentage, fill = Candidate)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.2f%%", Percentage)), vjust = -0.5, size = 3) +
  labs(y = "Percentage", x = "Candidate") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_classic() +
  theme(
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```

Additionally, [@fig-heatmap] presents a heat map of Harris's predicted probability by state, illustrating her lead over the majority of states but revealing that this advantage is generally weak, with most not peaking over a 1% advantage. Most states show only a slight preference for Harris over Trump, with the margin of error spanning both candidates in many cases. Specifically, out of the 43 states with available data, 25 states lean toward Harris, yet none of these states shows a substantial lead over more than a few percent over Trump. This pattern additionally shows the substantial closeness of the race, as evidenced by the margin of error, where overlapping intervals highlight the slim advantage and limited certainty in Harris's predicted support.
```{r}
#| label: fig-heatmap
#| fig-cap: "Heatmap of Predicted Support for Harris by State"
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 5
#| fig-height: 3

@@

# Plot heatmap for Harris's support by state, with NA values as dark grey
ggplot(merged_data, aes(x = long, y = lat, group = group, fill = avg_predicted_prob_harris)) +
  geom_polygon(color = "black", size = 0.3) + 
  scale_fill_gradient(low = "lightblue", high = "blue", na.value = "darkgrey", name = "Predicted Support") +
  coord_fixed(1.3) + 
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_void() +
  theme(
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  )
```

[@fig-swing-states] zooms in on the seven swing states, each of which shows a slight lean toward Harris. The graph includes a red dashed line at the 50% threshold, and although Harris is estimated to lead in these swing states, the margin is extremely tight, with most probabilities hovering just above 50%. Only Michigan and North Carolina show a slight lead of just over 1%, reinforcing the sensitivity of these predictions. This pattern suggests that even small shifts in voter sentiment could tip these critical states in either direction.
```{r}
#| label: fig-swing-states
#| fig-cap: "Predicted Probability of Harris Winning by State with Swing State Highlight and 50% Threshold"
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 6
#| fig-height: 4

@@

# Plot 
ggplot(state_predictions, aes(x = reorder(state, avg_predicted_prob_harris), y = avg_predicted_prob_harris, fill = swing_state)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  scale_fill_manual(values = c("Swing State" = "darkblue", "Non-Swing State" = "lightgrey")) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red", size = 0.8) +
  labs(
    x = "State",
    y = "Predicted Probability for Harris",
    fill = "State Type"
  ) +
  annotate("text", x = 1, y = 0.51, label = "0.5", color = "red", hjust = 0, size = 4) +
  theme_classic() +
  theme(
    axis.text.y = element_text(size = 5),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6)
  ) +
  coord_flip()
```
\newpage
Looking closer at the margin of error, in [@fig-margin] we can see Harris’s predicted probabilities by state. States to the right of the dashed line (above 0.5) lean towards Harris, while those to the left lean towards Trump. In swing states like North Carolina, Michigan, and Pennsylvania, Harris’s minimal lead is coupled with error bars that overlap this threshold. This suggests a high level of uncertainty—small shifts could easily change the lead in these states. Harris has a slight edge, but in many states, this lead is within the margin of error, thus showing that the election remains too close to call.
```{r}
#| label: fig-margin
#| fig-cap: "State-Level Predicted Probability for Harris with Margin of Error"
#| echo: false
#| warning: false
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
# Calculate margin of error for each state
state_margins <- clean_president_polls %>%
  group_by(state) %>%
  summarize(
    avg_predicted_prob_harris = mean(predicted_prob_harris),
    se_predicted_prob_harris = sd(predicted_prob_harris) / sqrt(n()),
    margin_of_error = 1.96 * se_predicted_prob_harris
  ) %>%
  mutate(
    lower_bound = avg_predicted_prob_harris - margin_of_error,
    upper_bound = avg_predicted_prob_harris + margin_of_error
  )

# Plot with error bars
ggplot(state_margins, aes(x = reorder(state, avg_predicted_prob_harris), y = avg_predicted_prob_harris)) +
  geom_point(color = "blue") +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2, color = "blue") +
  labs(
    x = "State",
    y = "Predicted Probability for Harris"
  ) +
  coord_flip() +
  theme_classic() +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  labs(subtitle = "Dashed line represents a 50% threshold") +
  theme(plot.subtitle = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 4),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8)
    ) 
```

@@

```{r}
#| label: tbl-summary-mod
#| tbl-cap: "Summary of Bayesian Model"
#| echo: false
#| warning: false
#| tbl-align: center
#| fig-width: 6
#| fig-height: 4

@@

# Appendix A: Methodology Analysis of The Washington Post Polling  {#sec-appenxA}
With an evaluation of sampling methodology, recruitment, handling non-response, and questionnaire design, this appendix offers an analysis of the polling methodology used by The Washington Post in collaboration with ABC News. The objective is to analyze these approaches' strengths and weaknesses and determine how they affect polling accuracy.

@@

# Appendix B: Idealized Survey & Methodology - $100K Budget {#sec-appenxB}

@@

This survey methodology uses stratified sampling, multi-channel recruitment, and data validation procedures to ensure accurate forecasting of the 2024 US Presidential Election. Through poll-of-polls aggregation, it offers a stable and reliable prediction by mitigating fluctuations in individual polls. The design aims to balance accuracy, inclusivity, and efficiency within a $100K budget, offering meaningful insights into voter behavior and sentiment. 
\newpage

# Appendix C {#sec-appc}
* [@tbl-pp]
```{r}
#| include: false
#| warning: false
#| message: false
### Model Validation/Checking ###
# Set seed and create train/test split
set.seed(123)
train_indices <- sample(seq_len(nrow(clean_president_polls)), size = 0.7 * nrow(clean_president_polls))
training_data <- clean_president_polls[train_indices, ]
training_data <- training_data %>%
  mutate(pct_scaled = scale(pct))
testing_data <- clean_president_polls[-train_indices, ]
